---
title: "DATA SOURCE: Case study on Soil Moisture"
description: "An example from the Awsome Public Datasets list"
weight: 38
categories:
- real_data
date: "2019-05-30"
---

```{r include = FALSE}
library(dplyr)
library(ggformula)
library(randomForest)
```
 
The website [Awesome public datasets](https://github.com/awesomedata/awesome-public-datasets#readme) contains links to several hundred datasets.  This example deals with  a randomly selected one, [Hyperspectral benchmark dataset on soil moisture](https://zenodo.org/record/1227837#.XORqSlNKjOR)

That dataset is described as being a sequence of measurements of a soil sample, recording temperature, soil moisture, and the intensity at each of 125 light frequency bands (in nm) of  an image of the sample.

## Process

1. Download the ZIP file.
2. Uncompress the ZIP file

This results in a  directory with a CSV file, a README.txt, and a license file.

3. Use the "Import Dataset" tool in the Environment tab to read in  the data. The tool constructs a command, which is used in the following chunk.
```{r}
library(readr)
Soil <- 
   read_csv("~/Downloads/felixriese-hyperspectral-soilmoisture-dataset-657d5d0/soilmoisture_dataset.csv", 
   col_types = cols(datetime = col_datetime(format = "%Y-%m-%d  %H:%M:%S")))
```

Notice the datetime format string used in the command. If working with time data,  it's helpful to know about how datetimes can be converted to machine readable  format.

Looking at the `Soil` data  frame ...

```{r}
nrow(Soil)
head(names(Soil))
range(Soil$datetime)
```

So we  have slightly more  than 10 days of measurements.

## What  to use the data for

The data were not gathered specifically for the  purpose of teaching statistics, so there is no documentation  that we can draw on to decide how to  fit  this  dataset  into  a  course. Creativity is required. Some ideas:

1. Telling a story with simple graphics. Look at the time series of soil temperature  and of soil_moisture. 
     a. How much of  moisture is accounted for by temperature.
    ```{r out.height = 2}
library(ggformula)
library(dplyr)
gf_point(soil_temperature ~ datetime,  data = Soil)
gf_point(soil_moisture ~ datetime, data = Soil) 
```
    b. What time of day were the measurements taken?
    ```{r}
    Soil %>%
      mutate(time = lubridate::hour(datetime) + lubridate::minute(datetime) / 60) %>%
      gf_jitter(time ~  1, data = Soil, width = 0.2) %>%
      gf_violin(fill  = "blue", alpha = 0.4,  color = NA) %>%
      gf_lims(x  = c(0, 2))
```
    c. Soil moisture as a function of time of day
    ```{r}
    Soil %>% 
      mutate(time = lubridate::hour(datetime) + lubridate::minute(datetime) / 60,
             day = as.character(lubridate::mday(datetime))) %>%
      gf_point(soil_moisture ~  time, color = ~  day, data = Soil) 
    ```
    
2. Simple analysis. Is soil moisture a function  of temperature?
    a. Is there a correlation?
    ```{r}
gf_point(soil_moisture ~ soil_temperature, data = Soil)
```
    b. Illuminating the pattern.
    ```{r}
    Soil %>%
      mutate(time = lubridate::hour(datetime) + lubridate::minute(datetime) / 60,
             day = as.character(lubridate::mday(datetime))) %>%
      gf_path(soil_moisture ~ soil_temperature, color = ~ day)
    ```
    
## More advanced?

Perhaps we can use the spectral measures  to  read soil moisture?

```{r message=FALSE}
library(rpart)
library(rpart.plot)
library(randomForest)
# need to rename variable
newnames <- function(df) {
  orig <- names(df)
  new <- ifelse(is.na(parse_number(orig)),  orig, paste0("v",  orig))
  return(new)
}
names(Soil) <- newnames(Soil)
mod1 <-  
  Soil %>%
  select(-index, -datetime) %>%
  randomForest(soil_moisture ~ . - soil_temperature, data = .)
Tmp <- importance(mod1) 
Res <- tibble(score = Tmp[, 1], wavelength = row.names(Tmp)) %>%
  arrange(desc(score))
head(Res)
gf_point(soil_moisture ~ v950,  data = Soil)
gf_point(v950 ~ v454,  color = ~  soil_moisture, data = Soil)
```