---
title: "DATA SOURCE: Big data from machine-learning sites"
description: "Records from a YouTube experiment to compare comedy videos"
weight: 38
author: "Danny Kaplan"
categories:
- real_data
date: "2019-05-30"
---

```{r include=FALSE}
library(dplyr)
library(ggformula)
```

One source of fairly large datasets  is machine-learning repositories or  contests such as  Kaggle.

This example concerns a dataset available at a [repository from the University of California at Irvine](https://archive.ics.uci.edu/ml/datasets.php).

The data set, [available here](https://archive.ics.uci.edu/ml/datasets/YouTube+Comedy+Slam+Preference+Data#), is about YouTube viewer's preferences among comedy videos. Attractive features of the dataset are:

1. It has a simple structure: just three variables.
2. The videos are (often?) still available on YouTube, providing some motivation for working with the data.

## Process

1. Download the ZIP file from the repository
2. UNZIP
3. Read in  the data

```{r message = FALSE}
file_name <- "/Users/kaplan/Downloads/comedy_comparisons/comedy_comparisons.train"
Clips <- readr::read_csv(file_name, col_names = FALSE)
names(Clips) <- c("left", "right", "winner")
```

## What to do

As usual, it isn't clear what statistically informative concept can be illustrated with these data. It's necessary to explore and see where the data takes you.

In the end, I decided on  a simple-sounding question: Which is the best clip?  The answer, as you'll see, involves balancing the win rate and the number of trials.

Here's the winner ...

<iframe width="280" height="157" src="https://www.youtube.com/embed/HE7cxOa9pXg" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

## Exploring

The data are organized as three variables:

`left` and `right` contain ID numbers for videos on YouTube.  `winner` indicates whether the left video or the right video were preferred by the viewer.

Some questions:

1. How many reviews are there? How many videos?
    ```{r}
    nrow(Clips)
    length(unique(c(Clips$left, Clips$right)))
    ```

2. Which is the most preferred video?
    ```{r}
    Wins <- 
      Clips %>%
      mutate(video = ifelse(winner == "left", left, right)) %>%
      group_by(video) %>%
      summarize(wins  = n()) %>%
      arrange(desc(wins)) 
    head(Wins)
    tail(Wins)
    ```

But see the  section on win rate below.

3. How many contests did each video appear in?
    ```{r}
    Lefts <- Clips %>% ungroup() %>% group_by(left) %>% summarize(nleft = n())
    Rights <- Clips %>% ungroup() %>% group_by(right) %>% summarize(nright = n()) 
    Contests <-  Lefts %>%
      inner_join(Rights, by = c(left = "right")) %>%
      mutate(total  =  nleft + nright) %>%
      select(left, total)
```

4. What's the win rate? Perhaps  insist that the p-value  against  the null of prob one-half be smaller than 0.001 or 0.0001. Or look at the lower bound of the 95% confidence interval on the number of wins.
    ```{r}
    Scores <- Wins %>% inner_join(Contests, by = c(video = "left")) %>%
      mutate(win_rate = wins / total) %>%
      mutate(p_value = 1 - pbinom(wins-1, p = 0.5, size = total )) %>%
      mutate(lower = qbinom(.025,  total, win_rate)) %>%
      mutate(pessimist = lower / total)
    ```
      
Can you pick which is the best clip?      
```{r cache = TRUE}
Plot <- gf_point(win_rate ~ log10(total), data  = Scores, alpha  = 0.03)
Plot
```

The highest scoring clips are among those with the fewest views.


According to the  p-value, which is the best movie?

```{r}
Scores %>%  
  filter(p_value < 0.0001) %>%
  arrange(desc(pessimist))
```

Links to be best videos ...

Win rates: [92%](https://www.youtube.com/watch?v=HE7cxOa9pXg), [80%](https://www.youtube.com/watch?v=hED4Tz1JM50)

