---
title: "DATA SOURCE: Hospital infections"
weight: 36
description: "Medicare hospital-level data on infection."
categories:
- real_data
date: "2019-05-30"
---

```{r include=FALSE}
library(dplyr)
library(ggformula)
library(readr)
```

## Medicare data

<data.medicare.gov/Hospital-Compare>: A web site suggested by Kim Druschel at Saint Louis University

One dataset there is about infections associated with hospitalization.
<https://data.medicare.gov/Hospital-Compare/Healthcare-Associated-Infections-Hospital/77hc-ibv8>

1. Download as CSV
2. In RStudio, go to Environment/ImportDataset

## Wrangling data:

```{r}
library(dplyr, quietly = TRUE, verbose = FALSE)
library(readr, quietly = TRUE, verbose = FALSE)
Infections <- readr::read_csv("~/Downloads/Healthcare_Associated_Infections_-_Hospital.csv")
unique(Infections$`Measure Name`)
# separate institution data from measurements
measurements <- names(Infections)[9:15]
Institutions <-
  Infections %>%
  select( - one_of(measurements)) %>%
  group_by(`Provider ID`) %>%
  filter(row_number() == 1)
Measures <- 
  Infections %>%
  select(`Provider ID`, one_of(measurements))
```

Let's look at central line associated bloodstream infections.
```{r}
Central_Line <- 
  Measures %>%
  filter(grepl("Central Line Associated", `Measure Name`)) %>%
  mutate(Score = ifelse(Score %in% c("--", "Not Available"), NA, parse_number(Score))) %>%
  mutate(condition = gsub("HAI_._", "", `Measure ID`)) %>%
  select(-`Measure Start Date`, -`Measure End Date`,  -`Footnote`, 
         -`Compared to National`, -`Measure Name`, -`Measure ID`) %>%
  tidyr::spread(key = condition, value = Score)
```

```{r}
Abdominal <- 
  Measures %>%
  filter(grepl("SSI - Abdom", `Measure Name`)) %>%
  mutate(Score = ifelse(Score %in% c("--", "Not Available"), NA, parse_number(Score))) %>%
  mutate(condition = gsub("HAI_._", "", `Measure ID`)) %>%
  select(-`Measure Start Date`, -`Measure End Date`,  -`Footnote`, 
         -`Compared to National`, -`Measure Name`, -`Measure ID`) %>%
  tidyr::spread(key = condition, value = Score)
Colon <- 
  Measures %>%
  filter(grepl("SSI - Colon", `Measure Name`)) %>%
  mutate(Score = ifelse(Score %in% c("--", "Not Available"), NA, parse_number(Score))) %>%
  mutate(condition = gsub("HAI_._", "", `Measure ID`)) %>%
  select(-`Measure Start Date`, -`Measure End Date`,  -`Footnote`, 
         -`Compared to National`, -`Measure Name`, -`Measure ID`) %>%
  tidyr::spread(key = condition, value = Score)
MRSA <- 
  Measures %>%
  filter(grepl("MRSA", `Measure Name`)) %>%
  mutate(Score = ifelse(Score %in% c("--", "Not Available"), NA, parse_number(Score))) %>%
  mutate(condition = gsub("HAI_._", "", `Measure ID`)) %>%
  select(-`Measure Start Date`, -`Measure End Date`,  -`Footnote`, 
         -`Compared to National`, -`Measure Name`, -`Measure ID`) %>%
  tidyr::spread(key = condition, value = Score)
Catheter <- 
  Measures %>%
  filter(grepl("Catheter", `Measure Name`)) %>%
  mutate(Score = ifelse(Score %in% c("--", "Not Available"), NA, parse_number(Score))) %>%
  mutate(condition = gsub("HAI_._", "", `Measure ID`)) %>%
  select(-`Measure Start Date`, -`Measure End Date`,  -`Footnote`, 
         -`Compared to National`, -`Measure Name`, -`Measure ID`) %>%
  tidyr::spread(key = condition, value = Score)
C_DIFF <- 
  Measures %>%
  filter(grepl("Clostridium", `Measure Name`)) %>%
  mutate(Score = ifelse(Score %in% c("--", "Not Available"), NA, parse_number(Score))) %>%
  mutate(condition = gsub("HAI_._", "", `Measure ID`)) %>%
  select(-`Measure Start Date`, -`Measure End Date`,  -`Footnote`, 
         -`Compared to National`, -`Measure Name`, -`Measure ID`) %>%
  tidyr::spread(key = condition, value = Score)

```


Can  we make sense of the variables?  Hypotheses ...

- `ELIGCASES` refers to a model output based on the number of patient days `DOPC`

```{r}
library(ggformula, quietly = TRUE, verbose = FALSE)
gf_point(ELIGCASES ~ DOPC, data = Central_Line)
gf_point(NUMERATOR ~ ELIGCASES, data = Central_Line) %>%
  gf_abline(intercept = 0, slope = 1,  color = "red")
gf_errorbar(CILOWER + CIUPPER ~ sqrt(ELIGCASES), data = Central_Line) 
```

CASE STUDY: Which providers to check out ...

* Is `ELIGCASES` too permissive?
* Which providers have too high an infection rate?
* Which providers have an admirably low infection rate?

```{r}
Too_high <- 
  Central_Line %>%
  filter(CILOWER > 1) %>%
  select(CILOWER, `Provider ID`) %>%
  arrange(desc(CILOWER)) 
Baddies <- 
  Institutions %>%
  inner_join(Too_high)
Nice <- 
  Central_Line %>%
  filter(CIUPPER < 1) %>%
  select(CIUPPER, `Provider ID`) %>%
  arrange(desc(CIUPPER)) 
Goodies <-
  Institutions %>%
  inner_join(Nice)
```

Activity about confidence intervals: Which under-performing hospitals should we  prioritize for inspection?  Use the confidence intervals being above 1 to identify them.