---
title: "Welcome"
description: "Outline of workshop and its goals"
categories:
  - presentations
date: "2019-06-07"
weight: 1
banner: "img/welcome.jpeg"
---

```{r   include=FALSE}
  library("NLP")
  library("tm")
  library("SnowballC")
  library("wordcloud")
  library("RColorBrewer") 
#++++++++++++++++++++++++++++++++++
# rquery.wordcloud() : Word cloud generator
# - http://www.sthda.com
#+++++++++++++++++++++++++++++++++++
# x : character string (plain text, web url, txt file path)
# type : specify whether x is a plain text, a web page url or a file path
# lang : the language of the text
# excludeWords : a vector of words to exclude from the text
# textStemming : reduces words to their root form
# colorPalette : the name of color palette taken from RColorBrewer package, 
  # or a color name, or a color code
# min.freq : words with frequency below min.freq will not be plotted
# max.words : Maximum number of words to be plotted. least frequent terms dropped
# value returned by the function : a list(tdm, freqTable)
rquery.wordcloud <- function(x, type=c("text", "url", "file"), 
                          lang="english", excludeWords=NULL, 
                          textStemming=FALSE,  colorPalette="Dark2",
                          min.freq=3, max.words=200)
{ 

  
  if(type[1]=="file") text <- readLines(x)
  else if(type[1]=="url") text <- html_to_text(x)
  else if(type[1]=="text") text <- x
  
  # Load the text as a corpus
  docs <- Corpus(VectorSource(text))
  # Convert the text to lower case
  docs <- tm_map(docs, content_transformer(tolower))
  # Remove numbers
  docs <- tm_map(docs, removeNumbers)
  # Remove stopwords for the language 
  docs <- tm_map(docs, removeWords, stopwords(lang))
  # Remove punctuations
  docs <- tm_map(docs, removePunctuation)
  # Eliminate extra white spaces
  docs <- tm_map(docs, stripWhitespace)
  # Remove your own stopwords
  if(!is.null(excludeWords)) 
    docs <- tm_map(docs, removeWords, excludeWords) 
  # Text stemming
  if(textStemming) docs <- tm_map(docs, stemDocument)
  # Create term-document matrix
  tdm <- TermDocumentMatrix(docs)
  m <- as.matrix(tdm)
  v <- sort(rowSums(m),decreasing=TRUE)
  d <- data.frame(word = names(v),freq=v)
  # check the color palette name 
  if(!colorPalette %in% rownames(brewer.pal.info)) colors = colorPalette
  else colors = brewer.pal(8, colorPalette) 
    # Plot the word cloud
  set.seed(1234)
  wordcloud(d$word,d$freq, min.freq=min.freq, max.words=max.words,
            random.order=FALSE, rot.per=0.35, 
            use.r.layout=FALSE, colors=colors)
  
  invisible(list(tdm=tdm, freqTable = d))
}
  #++++++++++++++++++++++
# Helper function
#++++++++++++++++++++++
# Download and parse webpage
html_to_text<-function(url){
  library(RCurl)
  library(XML)
  # download html
  html.doc <- getURL(url)  
  #convert to plain text
  doc = htmlParse(html.doc, asText=TRUE)
 # "//text()" returns all text outside of HTML tags.
 # We also don’t want text such as style and script codes
  text <- xpathSApply(doc, "//text()[not(ancestor::script)][not(ancestor::style)][not(ancestor::noscript)][not(ancestor::form)]", xmlValue)
  # Format text vector into one character string
  return(paste(text, collapse = " "))
}
```
  
## What are you  looking  for? 


```{r echo = FALSE,  warning  = FALSE, out.width = "100%"}
filePath <- "why.txt"
rquery.wordcloud(filePath, type ="file", lang = "english", textStemming = FALSE,
                      min.freq = 5)
```

## What challenges do you face?

```{r echo = FALSE,  warning  = FALSE, out.width = "100%"}
filePath <- "challenges.txt"
rquery.wordcloud(filePath, type ="file", lang = "english", textStemming = FALSE, min.freq = 5)
```

## What are institutional challenges

```{r echo = FALSE,  warning  = FALSE, out.width = "100%"}
filePath <- "institution.txt"
rquery.wordcloud(filePath, type ="file", lang = "english", textStemming = FALSE, min.freq = 5)
```

## Background and experience

From the pre-workshop survey

- Institution type
    - Two-year college 17/34
    - Four year institution 16/34
- Grad-school discipline
   - Mathematics  30 / 34
   - Statistics 9 / 34 
   - OR / Economics / Healthcare Admin 3 / 34
- Statistics experience
   - studied applied statistics in  grad school 18 / 31
   - studied theoretical statistics in grad school 15 / 33
   - do applied statistics outside of teaching 10 / 34
- Teaching experience: number with > 2 years experience
    - "other" mathematics  26 / 31
        - intro stat 23 / 32 (2 have no experience)
        - college algebra  21 / 31
    - calculus (1 has no experience) 19 / 31
    - non-intro statistics (7 have  no experience) 13 / 31
 
### Changing faster and faster ... 

How has the content and application of our field changed?

* Statistics:
     
    - Bayes and MCMC (since 1990)
    - Databases (since 1965)
    - Confounders and causation (since 1960)
    - Modern modes of graphics (since approx. 1990)
    - Machine learning (since 2010)

* College algebra:

    - ?
    
* Calculus

    - ?
    
* Trigonometry

    - ?
    
For examples of change, see [Nick Horton's](http://www.math.smith.edu/~nhorton/doctor.pdf) article on changes in the *New England Journal of Medicine* over the 15 years to 2005.
    
### How teaching calculus might distort a view of statistics

Example of calculus-like problem ... of a sort that I very rarely see in statistical work.

*The scores on the SAT verbal test in recent years follow approximately the N(517, 112) distribution.*

- What is the proportion of students scoring under 400? 
- What is the proportion of students scoring between 400 and 550? 
- How high must a student score to place in the top 10% of all students taking the SAT? State answer as a whole number. 
- Using the empirical rule, what is the probability that a randomly [selected] SAT test will have a verbal score between 629 and 853?
    
   
### Goals countdown: 3, 2, 1, ...

- **GAISE Recommendation 3**: Integrate real data with a context and a purpose.
    - You: 30 / 33 said learning how to incorporate real data into classes is their important or their top priority. 12 / 33 said "top priority" (So we'll have to talk about what is "real data.")

- **GAISE Recommendation 2**: Focus on conceptual understanding.
    * "*Pare down content of an introductory course to focus on core concepts in more depth.*"
    - You: 29 / 32 -- The many methods covered in introductory statistics can be reduced to a small set of common principles.
        * What are those principles? [Write them down here](https://docs.google.com/document/d/12rFpIZnr6abpkpfbRD-GFixI3T-vuX8zvYAUd-2WBu4/edit?usp=sharing)
    
    * "*Perform most computations using technology to allow greater emphasis on understanding concepts and interpreting results.*"

    - You: 29 / 32 Computing offers a framework for understanding statistical theory that is as legitimate as the theory based on probability rules and algebra.
    - You: Only 8 / 33 were confident in using computer programs to analyze data or data sets
    - You: 30 / 33 Proficiency in using computers to handle and manage data should be an important goal of a statistics course.

- **GAISE Recommendation 1**: Teach statistical thinking.

    - Teach statistics as an investigative process of problem-solving and decision-making
        * *"[I]nstructors [should] illustrate the complete investigative cycle with every example/exercise presented."*

    - Give students experience with multivariable thinking.
        - Characterizing relationships involving multiple (more than 3) variables:
            Never: 12/30, Rarely 9/30, Sometimes 9/30
        - You: Only 9 / 33 were confident in **developing data models**.

## Our goals for you ...

We hope you will leave this workshop able and interested to use real data to teach the topics you currently cover.

Ultimately, embracing real data in the ways needed in the workplace and as envisioned by GAISE is going to call for broader changes:

- Streamlining the topics/methods of your course
- Becoming a role model for proper management of data
- Confronting statistical issue that are important to contemporary uses of data but which are not part of the historic canon of statistics:
    - prediction
    - decision-making
    - causality






## Three components of the workshop

1. Things you can directly use in teaching:
    - Exploring data-driven activities via Little Apps
    - We've got about 20 activities on a variety of topics in statistics. 
2. Faculty development: Statistics topics the textbook doesn't cover
    - Bootstrapping
    - Unifying inference with regression
3. Faculty development: Data science
    - Sources of data, graphing data
    - R ecosystem
    
Levels will range from the easy to the aspirational.



### Your priorities

- Top priority
    - 12 / 33 learn how to incorporate real data into classes (30 / 33) YES*
    - 18 / 33 learn new statistics teaching methods (29 / 33) YES*
    - 13 / 33 learn how to engage students (31 / 33) WE PLAN and HOPE*
    - 10 / 33 learning more about data science concepts (19 / 33) YES*
    - 10 / 33 learn about tools/technology (28 / 33) YES*
    - 8 / 33 learning about materials, data sets, ... (21 / 33)
    - 5 / 33 learning to  use R (15 / 33) IF YOU WANT*
    - 4 / 33 get ideas about modifying department curriculum (19 / 33) WE THINK SO*

All together, the 33 respondents have 97 "top" priorities!  

✽ Will we cover this in the next two days?

### Confidence

- 8 / 33 were confident in using computer programs to analyze data or data sets
- Only 9 / 33 were confident in **developing data models**.

Possible issues: 

- what does "developing" mean?
- what does "models" mean?
- and, maybe, what does "data" mean?


    


### Software vs by hand

- 1 / 33 -- by hand
- 3 / 33 -- by hand and graphing calculator
- 10 / 33 -- by software
- 13 / 33 -- by hand and also by software
- 6 / 33 -- by graphing calculator

### Statistical theory

agree or strongly agree

- 6 / 29 -- algebraic formulas the best method to express statistics, yet ...
    - 14 / 31 -- statistical concepts are intrinsically based in algebraic notation
    - 32 / 33 -- Technology tools should be used to illustrate concepts
- 25 / 32 -- include theoretical probability distributions
- 29 / 32 -- The many methods covered in introductory statistics can be reduced to a small set of common principles.
    * What are those principles? [Write them down here](https://docs.google.com/document/d/12rFpIZnr6abpkpfbRD-GFixI3T-vuX8zvYAUd-2WBu4/edit?usp=sharing)

### Computing

- 29 / 32 Computing offers a framework for understanding statistical theory that is as legitimate as the theory based on probability rules and algebra.
    - but, earlier, 14 / 31 said "statistical concepts are intrinsically based in algebraic notation"
- 30 / 33 Proficiency in using computers to handle and manage data should be an important goal of a statistics course.
- Why aren't you using preferred software?
    - 19 / 31 I haven’t had time to explore software / technology beyond what I’m currently using.
    - 13 / 31 Takes too much time for students to learn software / technology
    - 21 / 31 Expense or Department or Access
    - 12 / 31 Not enough time to teach software and cover course material
- What do you use for software?
    - 10 / 29 Spreadsheets
    - 19 / 29 Graphing calculator
    - 7 / 29 StatKey / Stat Crunch
    - 14 / 29 Web apps
    - 3 / 29 R/RStudio
    - 3 / 29 Minitab/SPSS
    - 1 / 29 SAS/JMP
- What would you like to use for software?
    - 21 / 30 R/RStudio
    - 18 / 30 Spreadsheets
    - 11 / 30 Graphing calculators
    - 11 / 30 Web apps
    - 6 / 30 Minitab / SPSS
    - 4 / 30 SAS/JMP
    - 1 / 30 "I would prefer to use as little technology as possible"
    
### Graphics

Indicate the method of computing that you believe helps students learn introductory statistics best.

- 23 / 33 -- Graphics routinely drawn using statistical software
- 7 / 33 -- Graphics routinely drawn by hand

**Question**: Do you teach them the algorithms and programming steps for producing graphics or just have them go at it with the appropriate software function?



### Success

Out of 270 possible answers:

- 5 / 270 are very successful
- 60 / 270 are successful
- 92 / 270 are moderately successful
- 23 / 270 not successful

Grade: B- overall

### Limits to change

- 20 / 31 Limited personal time
- 19 / 31 Student characteristics (e.g., ability, interest, etc.)
- 19 / 31 Institutional constraints
- 15 / 31 Technology constraints





