---
title: "Welcome"
description: "Outline of workshop and its goals"
categories:
  - presentation
date: "2019-06-07"
weight: 1
---

```{r   include=FALSE}
  library("NLP")
  library("tm")
  library("SnowballC")
  library("wordcloud")
  library("RColorBrewer") 
#++++++++++++++++++++++++++++++++++
# rquery.wordcloud() : Word cloud generator
# - http://www.sthda.com
#+++++++++++++++++++++++++++++++++++
# x : character string (plain text, web url, txt file path)
# type : specify whether x is a plain text, a web page url or a file path
# lang : the language of the text
# excludeWords : a vector of words to exclude from the text
# textStemming : reduces words to their root form
# colorPalette : the name of color palette taken from RColorBrewer package, 
  # or a color name, or a color code
# min.freq : words with frequency below min.freq will not be plotted
# max.words : Maximum number of words to be plotted. least frequent terms dropped
# value returned by the function : a list(tdm, freqTable)
rquery.wordcloud <- function(x, type=c("text", "url", "file"), 
                          lang="english", excludeWords=NULL, 
                          textStemming=FALSE,  colorPalette="Dark2",
                          min.freq=3, max.words=200)
{ 

  
  if(type[1]=="file") text <- readLines(x)
  else if(type[1]=="url") text <- html_to_text(x)
  else if(type[1]=="text") text <- x
  
  # Load the text as a corpus
  docs <- Corpus(VectorSource(text))
  # Convert the text to lower case
  docs <- tm_map(docs, content_transformer(tolower))
  # Remove numbers
  docs <- tm_map(docs, removeNumbers)
  # Remove stopwords for the language 
  docs <- tm_map(docs, removeWords, stopwords(lang))
  # Remove punctuations
  docs <- tm_map(docs, removePunctuation)
  # Eliminate extra white spaces
  docs <- tm_map(docs, stripWhitespace)
  # Remove your own stopwords
  if(!is.null(excludeWords)) 
    docs <- tm_map(docs, removeWords, excludeWords) 
  # Text stemming
  if(textStemming) docs <- tm_map(docs, stemDocument)
  # Create term-document matrix
  tdm <- TermDocumentMatrix(docs)
  m <- as.matrix(tdm)
  v <- sort(rowSums(m),decreasing=TRUE)
  d <- data.frame(word = names(v),freq=v)
  # check the color palette name 
  if(!colorPalette %in% rownames(brewer.pal.info)) colors = colorPalette
  else colors = brewer.pal(8, colorPalette) 
    # Plot the word cloud
  set.seed(1234)
  wordcloud(d$word,d$freq, min.freq=min.freq, max.words=max.words,
            random.order=FALSE, rot.per=0.35, 
            use.r.layout=FALSE, colors=colors)
  
  invisible(list(tdm=tdm, freqTable = d))
}
  #++++++++++++++++++++++
# Helper function
#++++++++++++++++++++++
# Download and parse webpage
html_to_text<-function(url){
  library(RCurl)
  library(XML)
  # download html
  html.doc <- getURL(url)  
  #convert to plain text
  doc = htmlParse(html.doc, asText=TRUE)
 # "//text()" returns all text outside of HTML tags.
 # We also don’t want text such as style and script codes
  text <- xpathSApply(doc, "//text()[not(ancestor::script)][not(ancestor::style)][not(ancestor::noscript)][not(ancestor::form)]", xmlValue)
  # Format text vector into one character string
  return(paste(text, collapse = " "))
}
```
  
## What are you  looking  for? 


```{r echo = FALSE,  warning  = FALSE, out.width = "100%"}
filePath <- "why.txt"
rquery.wordcloud(filePath, type ="file", lang = "english", textStemming = FALSE,
                      min.freq = 5)
```

## What challenges do you face?

```{r echo = FALSE,  warning  = FALSE, out.width = "100%"}
filePath <- "challenges.txt"
rquery.wordcloud(filePath, type ="file", lang = "english", textStemming = FALSE, min.freq = 5)
```

## What are institutional challenges

```{r echo = FALSE,  warning  = FALSE, out.width = "100%"}
filePath <- "institution.txt"
rquery.wordcloud(filePath, type ="file", lang = "english", textStemming = FALSE, min.freq = 5)
```



## Background and experience

From the pre-workshop survey

- Institution type
    - 8 / 19 Two-year college
    - 12 / 19 Four year institution
- Grad-school discipline
   - 18 / 19 Mathematics  (and maybe it's 19 / 19)
   - 2 / 19 Computer science
   - 2 / 19 Engineering
- Statistics experience
   - 3 / 19  studied applied statistics in  grad school
   - 2 / 18  studied theoretical statistics in grad school
   - 7 / 18  do applied statistics outside of teaching
- Teaching experience: number with > 2 years experience
    - 15 / 16 "other" mathematics (1 has no experience)
    - 14 / 19 calculus (1 has no experience)
    - 10 / 19 college algebra (4 have no experienceª)
    - 7 / 19 intro stat 
    - 3 / 16 computer science (12 have no experience)
    - 3 / 18 statistics (7 have  no experience)
   
✽ Might these be 4-year instructors at institutions without a college algebra course?

### How teaching calculus might distort a view of statistics

Example of calculus-like problem ... of a sort that I very rarely see in statistical work.

*The scores on the SAT verbal test in recent years follow approximately the N(517, 112) distribution.*

- What is the proportion of students scoring under 400? 
- What is the proportion of students scoring between 400 and 550? 
- How high must a student score to place in the top 10% of all students taking the SAT? State answer as a whole number. 
- Using the empirical rule, what is the probability that a randomly [selected] SAT test will have a verbal score between 629 and 853?


### Your priorities

- Top priority
    - 8 / 14 learn how to incorporate real data into classes (17  / 19) YES*
    - 7 / 14 learn about tools/technology (18 / 19) YES*
    - 7 / 14 learn new statistics teaching methods (18 / 19) YES*
    - 7 / 14 learn how to engage students (15 / 19) WE PLAN and HOPE*
    - 6 / 14 learning more about data science concepts (17 / 19) YES*
    - 6 / 14 learning to  use R (14 / 19) IF YOU WANT*
    - 5 / 14 get ideas about modifying department curriculum (11 / 19) WE THINK SO*

All together, the 14 respondents have 58 "top" priorities!  

✽ Will we cover this in the next two days?

### Confidence

Only 4 / 19 were confident in **developing data models**.

Possible issues: 

- what does "developing" mean?
- what does "models" mean?
- and, maybe, what does "data" mean?

### Software vs by hand

- 1 / 19 -- by hand
- 9 / 19 -- by software
- 9 / 19 -- by hand and also by software
- Nobody -- by graphing calculator

### Statistical theory

- 3 / 19 -- algebra is the best method to express statistics, yet ...
    - 10 / 18 -- statistical concepts are intrinsically based in algebraic notation
- 5 / 19 -- include theoretical probability distributions
- EVERYONE -- The many methods covered in introductory statistics can be reduced to a small set of common principles.
    * What are those principles? [Write them down here](https://docs.google.com/document/d/12rFpIZnr6abpkpfbRD-GFixI3T-vuX8zvYAUd-2WBu4/edit?usp=sharing)

### Computing

- 16 / 19 Computing offers a framework for understanding statistical theory that is as legitimate as the theory based on probability rules and algebra.
    - but, earlier, 10 / 18 said "statistical concepts are intrinsically based in algebraic notation"
- 18 / 19 Proficiency in using computers to handle and manage data should be an important goal of a statistics course.
- Why aren't you using preferred software?
    - 9 / 15 I haven’t had time to explore software / technology beyond what I’m currently using.
    - 4 / 15 Takes too much time for students to learn software / technology
    - Expense or Department or Access: at most 1 or 2 out of 15
- What do you use for software?
    - 9 / 14 Spreadsheets
    - 5 / 14 Graphing calculator
    - 4 / 14 Other
    - 3 / 14 Web apps
    - 2 / 14 R/RStudio
    - 1 / 14 Minitab/SPSS
    - 0 / 14 SAS/JMP
- What would you like to use for software?
    - 7 / 11 R/RStudio
    - 6 / 11 Spreadsheets
    - 4 / 11 SAS/JMP
    
### Graphics

This was the strongest consensus ...

Indicate the method of computing that you believe helps students learn introductory statistics best.

- 17 / 18 -- Graphics routinely drawn using statistical software

**Question**: Do you teach them the algorithms and programming steps for producing graphics or just have them go at it with the appropriate software function?



### Success

Out of 90 possible answers:

- 4 / 90 are very successful
- 36 / 90 are successful
- 51 / 90 are moderately successful
- 10 / 90 not successful

Grade: B- overall

### Limits to change

- 10 / 15 Limited personal time
- 11 / 15 Student characteristics (e.g., ability, interest, etc.)
- Others (institutional constraints, transferability, no access to computing) got at most 3/15 each

## Three components of the workshop

1. Things you can directly use in teaching:
    - Exploring data-driven activities via Little Apps
    - We've got about 20 activities on a variety of topics in statistics. 
2. Faculty development: Statistics topics the textbook doesn't cover
    - Bootstrapping
    - Unifying inference with regression
3. Faculty development: Data science
    - Sources of data, graphing data
    - R ecosystem
    
Levels will range from the easy to the aspirational.



